Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-22 methods.tex
more editing to code
cleaning up after meeting

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units, a clock rate of 3.40GHz, and a cache size of
8192 KB.  We gathered performance traces during the execution of three
different programs---the simple \col loop whose performance is
depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.  To record these measurements, we used the {\tt libpfm4}
library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.  They can be programmed to measure a
  wide array of processor and memory metrics besides the one used
  here.}, which we had programmed to count how many instructions were
executed in each clock cycle (IPC).  For an in-depth description of
this custom-measurement infrastructure, including discussion of the
implications of the sampling interval, please see
\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
we collected 15 performance traces from each of the three programs.
These traces, and the processes that generated them, are described in
more depth in the rest of this section.

@@

\col is a simple C program that repeatedly initializes the upper
triangle of a 2048 $\times$ 2048 matrix in column-major order by
looping over the following three lines of code:
\begin{lstlisting}
% #define SIZE (2048)
% 
% int main(int argc, char ** argv)
% {
%   int i, j, n;
%   static int data[SIZE][SIZE];
%   
%   for(n=0;n<numInits;n++){  
  for(i=0;i<SIZE;i++)
  	for(j=i;j<SIZE;j++)
  		data[j][i] = 0;
%   }
%   return 0;
}
\end{lstlisting}
\noindent As mentioned in Section~\ref{sec:intro}, this simple program
exhibits surprisingly complicated behavior~\cite{mytkowicz09}.  A time
series of the processor performance during the execution of this
program is shown in Figure~\ref{fig:col-ts}.
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/colFullTS}
    \caption{The instructions executed per CPU clock cycle (IPC) as
      the \col program runs.  Each point is the average IPC in a
      100,000 instruction period. Figure~\ref{fig:col-ipc} is a
      closeup of the first part of this signal.  }
    \label{fig:col-ts}

@@

The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
complicated programs that are used in the computer science community
to assess and compare the performance of different computers.  \gcc is
a member of that suite.
% 
% This program is written by Richard Stallman and is based on Version
% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
% many of its optimization flags enabled, compiling a series of input
% files and generating x86-64 assembly code files intended for execution
% on AMD Opteron processors\cite{spec}. 
% 
It is a \emph{compiler}: a program that translates code written in the
C language into a lower-level format that can be executed by the
processor chip.  Its behavior is far more complicated than that of
\col, as is clear from Figure~\ref{fig:gcc-ts}.
  \begin{figure}[t]
  \centering
    \includegraphics[width=\columnwidth]{figs/gccfullts}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \gcc. Each point is the average IPC in a 100,000
      instruction period.}
    \label{fig:gcc-ts}
  \end{figure}
Unlike \col, where the processor utilization is quite structured,
\gcc's performance appears almost random.

>>>>

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units, a clock rate of 3.40 GHz, and a cache size of
8192 KB.  We gathered performance traces during the execution of three
different programs---the simple \col loop whose performance is
depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.  To record these measurements, we used the {\tt libpfm4}
library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.  They can be programmed to measure a
  wide array of processor and memory metrics besides the one used
  here.}, which we had programmed to count how many instructions were
executed in each clock cycle (IPC).  For an in-depth description of
this custom-measurement infrastructure, including discussion of the
implications of the sampling interval, please see
\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
we collected 15 performance traces from each of the three programs.
These traces, and the processes that generated them, are described in
more depth in the rest of this section.

@@

\col is a simple C program that repeatedly initializes the upper
triangle of a 2048 $\times$ 2048 matrix in column-major order by
looping over the following three lines of code:
\begin{lstlisting}
  for(i=0;i<2048;i++)
  	for(j=i;j<2048;j++)
  		data[j][i] = 0;
\end{lstlisting}
% full listing
% #define SIZE (2048)
% 
% int main(int argc, char ** argv)
% {
%   int i, j, n;
%   static int data[SIZE][SIZE];
%   
%   for(n=0;n<numInits;n++){  
%   for(i=0;i<SIZE;i++)
%   	for(j=i;j<SIZE;j++)
%   		data[j][i] = 0;
%   }
%   return 0;
% }
%
\noindent As mentioned in Section~\ref{sec:intro}, this simple program
exhibits surprisingly complicated behavior~\cite{mytkowicz09}.  A time
series of the processor performance during the execution of this
program is shown in Figure~\ref{fig:col-ts}.
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/colFullTS}
    \caption{The instructions executed per CPU clock cycle (IPC) as
      the \col program runs.  Each point is the average IPC in a
      100,000 instruction period. Figure~\ref{fig:col-ipc} is a
      closeup of the first part of this signal.  }
    \label{fig:col-ts}

@@

The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
complicated programs that are used in the computer science community
to assess and compare the performance of different computers.  \gcc is
a member of that suite.
% 
% This program is written by Richard Stallman and is based on Version
% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
% many of its optimization flags enabled, compiling a series of input
% files and generating x86-64 assembly code files intended for execution
% on AMD Opteron processors\cite{spec}. 
% 
It is a \emph{C-compiler}: a program that translates code written in the
C language into a lower-level format that can be executed by the
processor chip.  Its behavior is far more complicated than that of
\col, as is clear from Figure~\ref{fig:gcc-ts}.
  \begin{figure}[t]
  \centering
    \includegraphics[width=\columnwidth]{figs/gccfullts}
    \caption{The instructions per CPU clock cycle (IPC) during the
      execution of \gcc. Each point is the average IPC in a 100,000
      instruction period.}
    \label{fig:gcc-ts}
  \end{figure}
Unlike \col, where the processor utilization is quite structured,
\gcc's performance appears almost random.

### 2014-04-22 model.tex
changed table caption
cleaning up after meeting

\item The \emph{naive} method, which uses the mean of the
  observed signal as the forecast,

@@

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to the
original autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$p_i
= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
computed generally using either an ordinary least squares approach, or with the method of moments using the Yule-Walker equations.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
structure is present in the data, a \emph{seasonal ARIMA model}, which
adds a sampling operation that filters out periodicities, can be a
good choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this.  This procedure uses sophisticated methods---KPSS
unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximized likelihood
of the model fitted to the detrended data---to select good values for
the free parameters of the ARIMA model.

@@

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the average mutual information as an
estimate of $\tau$~\cite{fraser-swinney} and using the
false-near-neighbor method of~\cite{KBA92} to estimate $m$.  Plots of
the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
embedded following this procedure, are shown in
Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the ``Lorenz method of
analogues'' (LMA), which is essentially nearest-neighbor prediction in
the embedded\footnote{Lorenz's original formulation used the full
  system state space; the method was later extended to embedded
  dynamics by Kantz~\cite{kantz97}.}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides accurate forecasts when the generating process
is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA does),
it can handle both linear and nonlinear processes.  If the underlying
generating process is nondeterministic, however, it can perform
poorly.  In Figure~\ref{fig:embedding}(b), for instance, little
structure is visible, so LMA may not work well.  More structure
appears to be present in~Figure~\ref{fig:embedding}(c), but this
reconstruction also appears to contain some noise.  The question as to
how much structure is present in a reconstruction---and how much of
that structure can be captured and used by LMA---is apropos of the
central question treated in this paper.  It may be that \gcc has some
redundancy that LMA cannot exploit, or that the structure in \svdfive
is effectively obfuscated, from the standpoint of the LMA method, by
noise.  By quantifying the balance between normalized redundancy (that
is, predictive structure) and entropy for these real-valued time
series, as shown in Section~\ref{sec:results}, we can begin to answer
these questions.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error between the predictions and the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, which can generate arbitrary-length
predictions\footnote{although the accuracy of these predictions will
  degrade over prediction length, in the presence of positive Lyapunov
  exponents}, but the performance of the other three methods used here
will degrade severely if the associated models are not periodically
rebuilt.  In order to make the comparison fair, we used an iterative
one-step prediction schema {\sl for all four methods}.  This has the
slightly confusing effect of causing the ``test'' signal to be used
both to assess the accuracy of each model and for periodic refitting.

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

>>>>

\item The \emph{\naive} method, which uses the mean of the
  observed signal as the forecast,

@@

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that to
make predictions.  The roots of this approach date back to the
original autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$p_i
= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
computed generally using either an ordinary least squares approach, or with the method of moments using the Yule-Walker equations.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
structure is present in the data, a \emph{seasonal ARIMA model}, which
adds a sampling operation that filters out periodicities, can be a
good choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this.  This procedure uses sophisticated methods---KPSS
unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood
of the model fitted to the detrended data---to select good values for
the free parameters of the ARIMA model.

@@

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as an
estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.  Plots of
the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
embedded following this procedure, are shown in
Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the ``Lorenz method of
analogues'' (LMA), which is essentially nearest-neighbor prediction in
the embedded\footnote{Lorenz's original formulation used the full
  system state space; According to \cite{kantz97} this method was first extended to embedded dynamics by Pikovsky ~\cite{pikovsky86-sov}, but more closely related to the prediction work of Sugihara \& May in \cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides accurate forecasts when the generating process
is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA does),
it can handle both linear and nonlinear processes.  If the underlying
generating process is nondeterministic, however, it can perform
poorly.  In Figure~\ref{fig:embedding}(b), for instance, little
structure is visible, so LMA may not work well.  More structure
appears to be present in~Figure~\ref{fig:embedding}(c), but this
reconstruction also appears to contain some noise.  The question as to
how much structure is present in a reconstruction---and how much of
that structure can be captured and used by LMA---is apropos of the
central question treated in this paper.  It may be that \gcc has some
redundancy that LMA cannot exploit, or that the structure in \svdfive
is effectively obfuscated, from the standpoint of the LMA method, by
noise.  By quantifying the balance between normalized (that
is, predictive structure) and entropy for these real-valued time
series, as shown in Section~\ref{sec:results}, we can begin to answer
these questions.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error between the predictions and the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, which can generate arbitrary-length
predictions---although the accuracy of these predictions will
  degrade over prediction length, in the presence of positive Lyapunov
  exponents, but the performance of the other three methods used here
will degrade severely if the associated models are not periodically
rebuilt.  In order to make the comparison fair, we used an iterative
one-step prediction schema {\sl for all four methods}.  This has the
slightly confusing effect of causing the ``test'' signal to be used
both to assess the accuracy of each model and for periodic refitting.

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

### 2014-04-22 related.tex
red text edits
cleaning up after meeting

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories: model-based error analysis and
model-free information analysis \alert{citation?}.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith {\sl et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

@@

A practice known as deterministic vs. stochastic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global approaches to model-based error analysis.
The basic idea is to construct a series of local linear fits,
beginning with a few points and working up to to a global linear fit
that includes all known points, and then analyze how the average
out-of-sample error changes as a function of number of points in the
fit. The shape of such a ``DVS" graph indicates the amounts of determinism
and stochasticity are present in a time series.

>>>>

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith {\sl et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

@@

A practice known as deterministic vs. stochastic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global approaches to model-based error analysis.
The basic idea is to construct a series of local linear fits,
beginning with a few points and working up to a global linear fit
that includes all known points, and then analyze how the average
out-of-sample error changes as a function of number of points in the
fit. The shape of such a ``DVS" graph indicates the amounts of determinism
and stochasticity present in a time series.\alert{Liz Should we say what this has to do with predictability, realized this paragraph may seem unrelated to a casual reader}

### 2014-04-22 results.tex
cleaning up after meeting

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

>>>>

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic\cite{mytkowicz09}, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.
